Covariate analysis of MAYO reprocessed counts (with TMM normalisation)

### 2017-03-08 code/Rmd/MSSM_RNASeq_Reprocessed_TMM_BinnedCDR.Rmd
Minor bug fix

# RIBOSOMAL FRACTION
p[[4]] = ggplot(METADATA, aes(x = BrodmannArea.Dx, y = as.numeric(PCT_RIBOSOMAL_BASES), fill = BrodmannArea)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('Fraction of Ribosomal Bases') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# INTRONIC BASES
p[[5]] = ggplot(METADATA, aes(x = BrodmannArea.Dx, y = as.numeric(PCT_INTRONIC_BASES), fill = BrodmannArea)) + geom_boxplot()
p[[5]] = p[[5]] + ggtitle('Frction Intronic Bases') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# Mapped Reads
p[[6]] = ggplot(METADATA, aes(x = BrodmannArea.Dx, y = as.numeric(METADATA$PCT_INTERGENIC_BASES), fill = BrodmannArea)) + geom_boxplot()
p[[6]] = p[[6]] + ggtitle('Fraction Intergenic Bases') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

>>>>

# Ribosomal bases
p[[4]] = ggplot(METADATA, aes(x = BrodmannArea.Dx, y = as.numeric(PCT_RIBOSOMAL_BASES), fill = BrodmannArea)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('Fraction of Ribosomal Bases') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# Intronic bases
p[[5]] = ggplot(METADATA, aes(x = BrodmannArea.Dx, y = as.numeric(PCT_INTRONIC_BASES), fill = BrodmannArea)) + geom_boxplot()
p[[5]] = p[[5]] + ggtitle('Frction Intronic Bases') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# Intergenic bases
p[[6]] = ggplot(METADATA, aes(x = BrodmannArea.Dx, y = as.numeric(PCT_INTERGENIC_BASES), fill = BrodmannArea)) + geom_boxplot()
p[[6]] = p[[6]] + ggtitle('Fraction Intergenic Bases') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

### 2017-03-08 code/Rmd/Mayo_RNASeq_Reprocessed_TMM.Rmd
Minor bug fix

METADATA = rbindlist(list(METADATA_TC, METADATA_CE), use.names = T, fill = T) %>%
  as.data.frame
```

@@

```{r diagnosisPlots, fig.height=6, fig.width = 10}
# RIN
p = list()
p[[1]] = ggplot(METADATA, aes(x = BrainRegion.Diagnosis, y = RIN, fill = Tissue)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

@@

multiplot(plotlist = p, cols = 3)

@@

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
ContCovariates <- c("RIN", "RIN2", "AgeAtDeath","PMI")

@@

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)
```
Covariate correlation
```{r covariates.correlation, fig.width=8, fig.height=8}
COVARIATES.CORRELATION = CovariateAnalysis::getAssociationStatistics(COVARIATES, PVAL = 0.05)
ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.4, row.width=0.1)
```
### Gene Filtering
Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape = Source, color = Region))
p <- p + theme_bw() + theme(legend.position = "left") +scale_shape_manual(values = 1:2)
p <- p + facet_grid(Region~.)
# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
p
```
```{r decompse.normalise.data2, fig.height=9, fig.width=25, results='asis'}
# Tree based clustering
sampleTrees = flashClust::hclust(dist(t(VOOM.GENE_EXPRESSION$E)), method = "average")
tree = WGCNA::cutreeStatic(sampleTrees, cutHeight = 135, minSize=10)

@@

# Find significant covariates
adjustCovars = preAdjustedSigCovars$significantCovars
```
Significant covariates to adjust at FDR 0.1 are `r paste(gsub('_','\\\\_',adjustCovars), collapse= ',')`
```{r preAdjustedSigCovars.NULL.ALL, fig.width=25, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (iterative)
Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix.
```{r iterative.adjusted.voom.normalisation, results='asis'}
postAdjustCovars = c('BrainRegion.Diagnosis', 'Source', 'FLOWCELL', 'Gender')
randomCovariates = 'Donor_ID'

# Assign residual covariates
residualSigCovars = preAdjustedSigCovars
significantCovars = residualSigCovars$significantCovars
covariatesEffects = residualSigCovars$Effects.significantCovars
covariatesEffects = covariatesEffects[setdiff(significantCovars, c(postAdjustCovars, randomCovariates))]

@@

  # Residual covariates to choose from
  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars, randomCovariates))

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                COVARIATES,
                                                'all adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

  # Add postadjusted covariates (if any)
  significantCovars = residualSigCovars$significantCovars
  covariatesEffects = residualSigCovars$Effects.significantCovars
  covariatesEffects = covariatesEffects[setdiff(significantCovars,c(postAdjustCovars, randomCovariates))]

@@

tmp <- paste('Using following covariates in the final model:', modelStr, 'and Donor_ID as random effects')
```
`r tmp`
### Sanity Check
Residualise significant covariates using a linear model and find correlation between PCA of residuals with covariates
```{r residualSigCovars.ALL, fig.width=25, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
Adding "BrainRegion.Diagnosis" back to the residuals
```{r addVarsBack}
varsToAddBackIn <-  grep("BrainRegion.Diagnosis", colnames(FIT$coefficients), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  t(data.matrix(DM1$design[,varsToAddBackIn]) %*% t(FIT$coefficients[,varsToAddBackIn]))
```

@@

```{r get.gene.symbols}
# Define biomart object
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl", host = "dec2016.archive.ensembl.org")

# Get background gene ids
backgroundGenes = data.frame(gene.id = rownames(COUNT), ID = rownames(COUNT)) %>%
  tidyr::separate(ID, c('ensembl_gene_id','position'), sep = '\\.')

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                       mart = mart)

# Add gene names
RESIDUAL.GENE_EXPRESSION <- rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'gene.id') %>%
  left_join(backgroundGenes) %>% 
  left_join(Ensemble2HGNC) %>%
  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))

VOOM.ADJUSTED.GENE_EXPRESSION$E <- rownameToFirstColumn(VOOM.ADJUSTED.GENE_EXPRESSION$E, 'gene.id') %>%
  left_join(backgroundGenes) %>% 
  left_join(Ensemble2HGNC) %>%
  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=20}
# Fit contrast
contrast = makeContrasts(contrasts=c(
  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.Pathologic_Aging",
  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.PSP",
  "BrainRegion.DiagnosisCER.Pathologic_Aging-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisCER.Pathologic_Aging-BrainRegion.DiagnosisCER.PSP",
  "BrainRegion.DiagnosisCER.PSP-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.AD",
  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.Pathologic_Aging",
  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.PSP",
  "BrainRegion.DiagnosisTCX.Pathologic_Aging-BrainRegion.DiagnosisTCX.AD",
  "BrainRegion.DiagnosisTCX.Pathologic_Aging-BrainRegion.DiagnosisTCX.PSP",
  "BrainRegion.DiagnosisTCX.PSP-BrainRegion.DiagnosisTCX.AD"),
  levels = colnames(FIT$coefficients))
FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:12, function(i, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION){
  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1], confint = T)
}, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION)
names(DE) = gsub('BrainRegion.Diagnosis','',colnames(contrast))

DE1 = plyr::ldply(DE, .id = 'Comparison') %>%
  dplyr::mutate(Comparison1 = Comparison) %>%
  tidyr::separate(Comparison1, c('from.state','to.state'), sep = '-') %>%
  tidyr::separate(from.state, c('BrainRegion','reference'), sep = '\\.') %>%
  tidyr::separate(to.state, c('BrainRegion','against'), sep = '\\.') %>%
  tidyr::unite(Name, reference, against, sep = '-') %>%
  dplyr::mutate(Comparison = Name, Study = 'MAYO') %>%
  dplyr::rename(gene.id = genes, Region = BrainRegion) %>%
  dplyr::select(-Name) %>%
  left_join(backgroundGenes) %>%
  left_join(Ensemble2HGNC)

p <- ggplot(DE1, aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
p <- p + geom_hline(yintercept = -log10(0.05), color = 'red')
p <- p + geom_vline(xintercept = log2(1.2), color = 'red')
p <- p + geom_vline(xintercept = -log2(1.2), color = 'red')
p <- p + facet_grid(Region~.+Comparison)
p

tmp = DE1 %>% 
  dplyr::filter(adj.P.Val <= 0.05) %>%
  group_by(Region, Comparison) %>%
  dplyr::summarize(nDiffExpGenes = length(unique(gene.id))) %>%
  spread(Region, nDiffExpGenes)
kable(tmp)
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = 'UFL_Mayo_ISB Reprocessed RNASeq TMM', parentId = parentId)
annotations(CODE) = list(
  dataType = 'mRNA',
  dataSubType = 'geneExp',
  summaryLevel = 'gene',
  assay	 = 'Reprocessed RNAseq',

@@

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'Sampleid')
write.table(COVARIATES, file = 'MAYO_CBE_TCX_Covariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = File('MAYO_CBE_TCX_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store filtered counts
PROCESSED_COUNTS$filteredExprMatrix %>%
  as.matrix() %>% rownameToFirstColumn('gene.id') %>%
  write.table(file = 'MAYO_CBE_TCX_Counts.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('MAYO_CBE_TCX_Counts.tsv', name = 'Raw counts (filtered)', parentId = CODE$properties$id)
COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                     executed = thisFile, activityDescription = activityDescription)

# Store logCPM
write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file = 'MAYO_CBE_TCX_logCPM.tsv', sep = '\t', row.names=F, quote=F)
LCOUNT_OBJ = File('MAYO_CBE_TCX_logCPM.tsv', name = 'Raw counts (logCPM)', parentId = CODE$properties$id)
LCOUNT_OBJ = synStore(LCOUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)

# Store design matrix
write.table(DM1$design, file = 'MAYO_CBE_TCX_Design.tsv', sep = '\t', row.names=T, quote=F)
DM_OBJ = File('MAYO_CBE_TCX_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                  executed = thisFile, activityDescription = activityDescription)

# Store residual gene expression
write.table(RESIDUAL.GENE_EXPRESSION, file = 'MAYO_CBE_TCX_Expression.tsv', sep = '\t', row.names=F, quote=F)
EXP_OBJ = File('MAYO_CBE_TCX_Expression.tsv', name = 'Normalised Adjusted Residual Expression (Dx added)', 
               parentId = CODE$properties$id)
EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store differential expression results
write.table(DE1, file = 'MAYO_CBE_TCX_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('MAYO_CBE_TCX_DiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)

>>>>

# Get picard metrics from synapse
METADATA.PICARD_CE_ID <- 'syn8380392'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_CE_ID
METADATA.PICARD_CE_FPATHS <- untar(synGet(METADATA.PICARD_CE_ID)@filePath, list = FALSE, exdir = getwd())
METADATA.PICARD_CE_FPATHS <- untar(synGet(METADATA.PICARD_CE_ID)@filePath, list = TRUE, exdir = getwd())
METADATA.PICARD_CE_FPATHS = METADATA.PICARD_CE_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_CE_FPATHS)]
METADATA.PICARD_CE <- plyr::ldply(METADATA.PICARD_CE_FPATHS, read.csv)

METADATA.PICARD_CE$ID = gsub('picard/', '', METADATA.PICARD_CE_FPATHS) %>%
  gsub('/picard.CombinedMetrics.csv','',.) %>%
  sapply(function(x){str_split(x, '\\.')[[1]][1]})
colnames(METADATA.PICARD_CE) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD_CE))
colnames(METADATA.PICARD_CE) = gsub('RnaSeqMetrics__','',colnames(METADATA.PICARD_CE))

METADATA.PICARD_TC_ID <- 'syn8383233'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_TC_ID
METADATA.PICARD_TC_FPATHS <- untar(synGet(METADATA.PICARD_TC_ID)@filePath, list = FALSE, exdir = getwd())
METADATA.PICARD_TC_FPATHS <- untar(synGet(METADATA.PICARD_TC_ID)@filePath, list = TRUE, exdir = getwd())
METADATA.PICARD_TC_FPATHS = METADATA.PICARD_TC_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_TC_FPATHS)]
METADATA.PICARD_TC <- plyr::ldply(METADATA.PICARD_TC_FPATHS, read.csv)

METADATA.PICARD_TC$ID = gsub('picard/', '', METADATA.PICARD_TC_FPATHS) %>%
  gsub('/picard.CombinedMetrics.csv','',.) %>%
  sapply(function(x){str_split(x, '\\.')[[1]][1]}) %>%
  gsub('Aligned','',.)
colnames(METADATA.PICARD_TC) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD_TC))
colnames(METADATA.PICARD_TC) = gsub('RnaSeqMetrics__','',colnames(METADATA.PICARD_TC))

# Merge all metadata
METADATA = rbindlist(list(METADATA_TC, METADATA_CE), use.names = T, fill = T) %>%
  inner_join(rbindlist(list(METADATA.PICARD_TC, METADATA.PICARD_CE), use.names = T, fill = T)) %>%
  as.data.frame
```

@@

```{r diagnosisPlots, fig.height=14, fig.width = 8}
# RIN
p = list()
p[[1]] = ggplot(METADATA, aes(x = BrainRegion.Diagnosis, y = RIN, fill = Tissue)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

@@

# Ribosomal bases
p[[4]] = ggplot(METADATA, aes(x = BrainRegion.Diagnosis, y = as.numeric(PCT_RIBOSOMAL_BASES), fill = BrainRegion)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('Fraction of Ribosomal Bases') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# Intronic bases
p[[5]] = ggplot(METADATA, aes(x = BrainRegion.Diagnosis, y = as.numeric(PCT_INTRONIC_BASES), fill = BrainRegion)) + geom_boxplot()
p[[5]] = p[[5]] + ggtitle('Frction Intronic Bases') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

# Intergenic bases
p[[6]] = ggplot(METADATA, aes(x = BrainRegion.Diagnosis, y = as.numeric(PCT_INTERGENIC_BASES), fill = BrainRegion)) + geom_boxplot()
p[[6]] = p[[6]] + ggtitle('Fraction Intergenic Bases') + theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

multiplot(plotlist = p, cols = 2)

@@

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
ContCovariates <- c("RIN", "RIN2", "AgeAtDeath","PMI", 
                    "PCT_PF_READS", "PCT_PF_READS_ALIGNED", "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", 
                    "PCT_RIBOSOMAL_BASES", "PCT_USABLE_BASES")

@@

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)
```
Covariate correlation
```{r covariates.correlation, fig.width=8, fig.height=8}
COVARIATES.CORRELATION = CovariateAnalysis::getAssociationStatistics(COVARIATES, PVAL = 0.05)
draw(COVARIATES.CORRELATION$plot, heatmap_legend_side = 'left')
```
### Gene Filtering
Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape = Source, color = Region))
p <- p + theme_bw() + theme(legend.position = "left") +scale_shape_manual(values = 1:2)
p <- p + facet_grid(Region~., scale = 'free_y')
# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
p
```
```{r decompse.normalise.data2, fig.height=9, fig.width=25, results='asis'}
# Tree based clustering
sampleTrees = flashClust::hclust(dist(t(VOOM.GENE_EXPRESSION$E)), method = "average")
tree = WGCNA::cutreeStatic(sampleTrees, cutHeight = 135, minSize=10)

@@

# Find significant covariates
adjustCovars = preAdjustedSigCovars$significantCovars
```
Significant covariates to adjust at FDR 0.1 are `r paste(gsub('_','\\\\_',adjustCovars), collapse= ',')`
```{r preAdjustedSigCovars.NULL.ALL, fig.width=25, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (iterative)
Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix
```{r iterative.adjusted.voom.normalisation, results='asis'}
postAdjustCovars = c('Source', 'FLOWCELL', 'Gender')
randomCovariates = 'Donor_ID'

# Assign residual covariates
residualSigCovars = preAdjustedSigCovars
significantCovars = residualSigCovars$significantCovars
covariatesEffects = residualSigCovars$Effects.significantCovars
covariatesEffects = covariatesEffects[setdiff(significantCovars, c(postAdjustCovars, randomCovariates, 'BrainRegion.Diagnosis'))]

@@

  # Residual covariates to choose from
  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars, randomCovariates, 'BrainRegion.Diagnosis'))

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                COVARIATES[, residCovars, drop=F],
                                                'all adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

  # Add postadjusted covariates (if any)
  significantCovars = residualSigCovars$significantCovars
  covariatesEffects = residualSigCovars$Effects.significantCovars
  covariatesEffects = covariatesEffects[setdiff(significantCovars,c(postAdjustCovars, randomCovariates, 'BrainRegion.Diagnosis'))]

@@

tmp <- paste('Using following covariates in the final model:', modelStr, 'and Donor_ID as random effects')
```
`r tmp`
### Sanity Check
Residualise significant covariates using a linear model and find correlation between PCA of residuals with covariates
```{r residualSigCovars.ALL, fig.width=25, fig.height=12}
# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              COVARIATES,
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)

residualSigCovars[["PC_res"]][[2]]$plotData
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

@@

```{r get.gene.symbols, cache=TRUE, eval=T}
backgroundGenes = data.frame(gene.id = rownames(COUNT)) %>%
  dplyr::mutate(ensembl_gene_id = gene.id) %>%
  tidyr::separate(ensembl_gene_id, c('ensembl_gene_id', 'position'), sep = '\\.')

# Define biomart object
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "dec2016.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", 
                       values = backgroundGenes$ensembl_gene_id,
                       mart = mart)
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=20}
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, 'BrainRegion.Diagnosis', drop = F], Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

# Fit linear model with residual expression
FIT = lmFit(RESIDUAL.GENE_EXPRESSION, design = DESIGN$design)

# Fit contrast
contrast = makeContrasts(contrasts=c(
  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.Pathologic_Aging",
  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.PSP",
  "BrainRegion.DiagnosisCER.Pathologic_Aging-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisCER.Pathologic_Aging-BrainRegion.DiagnosisCER.PSP",
  "BrainRegion.DiagnosisCER.PSP-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.AD",
  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.Pathologic_Aging",
  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.PSP",
  "BrainRegion.DiagnosisTCX.Pathologic_Aging-BrainRegion.DiagnosisTCX.AD",
  "BrainRegion.DiagnosisTCX.Pathologic_Aging-BrainRegion.DiagnosisTCX.PSP",
  "BrainRegion.DiagnosisTCX.PSP-BrainRegion.DiagnosisTCX.AD"),
  levels = colnames(FIT$coefficients))
FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:12, function(i, FIT.CONTR, RESIDUAL.GENE_EXPRESSION){
  topTable(FIT.CONTR, coef=i, number = dim(RESIDUAL.GENE_EXPRESSION)[1], confint = T) %>%
    rownameToFirstColumn('genes')
}, FIT.CONTR, RESIDUAL.GENE_EXPRESSION)
names(DE) = gsub('BrainRegion.Diagnosis','',colnames(contrast))

DE1 = plyr::ldply(DE, .id = 'Comparison') %>%
  dplyr::mutate(Comparison1 = Comparison) %>%
  tidyr::separate(Comparison1, c('from.state','to.state'), sep = '-') %>%
  tidyr::separate(from.state, c('BrainRegion','reference'), sep = '\\.') %>%
  tidyr::separate(to.state, c('BrainRegion','against'), sep = '\\.') %>%
  tidyr::unite(Name, reference, against, sep = '-') %>%
  dplyr::mutate(Comparison = Name, Study = 'MAYO') %>%
  dplyr::rename(gene.id = genes, Region = BrainRegion) %>%
  dplyr::select(-Name) %>%
  left_join(backgroundGenes) %>%
  left_join(Ensemble2HGNC) %>%
  dplyr::mutate(Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction))
DE1$Direction[DE1$adj.P.Val > 0.05 | abs(DE1$logFC) < log2(1.2)] = 'NONE'

p = ggplot(DE1, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red'))
p = p + facet_grid(Region~.+Comparison, scales = 'fixed')
p

tmp = DE1 %>% 
  dplyr::filter(adj.P.Val <= 0.05) %>%
  group_by(Region, Comparison) %>%
  dplyr::summarize(nDiffExpGenes = length(unique(gene.id))) %>%
  spread(Region, nDiffExpGenes)
kable(tmp)
```
### Associate differential expression results with gc content and gene length
```{r associate.de}
geneLength.GCcontent = EDASeq::getGeneLengthAndGCContent(unique(DE1$ensembl_gene_id), 
                                                         org = 'hsa', 
                                                         mode = "biomart") %>%
  rownameToFirstColumn('ensembl_gene_id')

tmp = left_join(DE1, geneLength.GCcontent)

pl = list()
pl[[1]] = ggplot(tmp, aes(y = log10(length), x = abs(logFC))) + 
  geom_point() + 
  stat_smooth(method=glm, se = TRUE, na.rm = T)
pl[[2]] = ggplot(tmp, aes(y = gc, x = abs(logFC))) + 
  geom_point() + 
  stat_smooth(method=glm, se = TRUE, na.rm = T)
pl[[3]] = ggplot(tmp, aes(y = log10(length), x = adj.P.Val)) + 
  geom_point() + 
  stat_smooth(method=glm, se = TRUE, na.rm = T)
pl[[4]] = ggplot(tmp, aes(y = gc, x = adj.P.Val)) + 
  geom_point() + 
  stat_smooth(method=glm, se = TRUE, na.rm = T)

multiplot(plotlist = pl, cols = 2)
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Add gene names
RESIDUAL.GENE_EXPRESSION <- rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'gene.id') %>%
  left_join(backgroundGenes) %>% 
  left_join(Ensemble2HGNC) %>%
  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))

VOOM.ADJUSTED.GENE_EXPRESSION$E <- rownameToFirstColumn(VOOM.ADJUSTED.GENE_EXPRESSION$E, 'gene.id') %>%
  left_join(backgroundGenes) %>% 
  left_join(Ensemble2HGNC) %>%
  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))

CODE <- Folder(name = 'UFL_Mayo_ISB Reprocessed RNASeq TMM', parentId = parentId)
annotations(CODE) = list(
  dataType = 'mRNA',
  dataSubType = 'geneExp',
  summaryLevel = 'gene',
  assay	 = 'Reprocessed RNAseq',

@@

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'ID')
write.table(COVARIATES, file = 'MAYO_Covariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = File('MAYO_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
annotations(COV_OBJ) = annotations(CODE)
annotations(COV_OBJ)$dataSubType = 'covariates'
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store filtered counts
PROCESSED_COUNTS$filteredExprMatrix %>%
  as.matrix() %>% rownameToFirstColumn('gene.id') %>%
  write.table(file = 'MAYO_Counts.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('MAYO_Counts.tsv', name = 'Raw counts (filtered)', parentId = CODE$properties$id)
annotations(COUNT_OBJ) = annotations(CODE)
annotations(COUNT_OBJ)$dataSubType = 'filteredCounts'
COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                     executed = thisFile, activityDescription = activityDescription)

# Store logCPM
write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file = 'MAYO_logCPM.tsv', sep = '\t', row.names=F, quote=F)
LCOUNT_OBJ = File('MAYO_logCPM.tsv', name = 'Raw counts (logCPM)', parentId = CODE$properties$id)
annotations(LCOUNT_OBJ) = annotations(CODE)
annotations(LCOUNT_OBJ)$dataSubType = 'logcpm'
LCOUNT_OBJ = synStore(LCOUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)

# Store design matrix
write.table(DM1$design, file = 'MAYO_Design.tsv', sep = '\t', row.names=F, quote=F)
DM_OBJ = File('MAYO_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
annotations(DM_OBJ) = annotations(CODE)
annotations(DM_OBJ)$dataSubType = 'designMatrix'
DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                  executed = thisFile, activityDescription = activityDescription)

# Store residual gene expression
write.table(RESIDUAL.GENE_EXPRESSION, file = 'MAYO_Expression.tsv', sep = '\t', row.names=F, quote=F)
EXP_OBJ = File('MAYO_Expression.tsv', name = 'Normalised Adjusted Residual Expression (Dx added)', 
               parentId = CODE$properties$id)
annotations(EXP_OBJ) = annotations(CODE)
annotations(EXP_OBJ)$dataSubType = 'residualGeneExp'
EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store differential expression results
write.table(DE1, file = 'MAYO_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('MAYO_DiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
annotations(DEXP_OBJ) = annotations(CODE)
annotations(DEXP_OBJ)$dataSubType = 'diffExp'
DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)
